---
title: 'Reproducible Research: Peer Assessment 1'
author: "funkje"
output:
  html_document:
    keep_md: yes
  pdf_document: default
---

## Loading libraries

* Data.table is my weapon of choice for manipulating data frames because of its speed and functionality.
* Ggplot2 lets me quickly prodcue nice graphs.

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figures/')
```

```{r include=FALSE}
options(scipen = 1, digits = 2) # display 2 decimal numbers for the sake of the report
```

```{r echo=TRUE}
library(data.table)
library(ggplot2)
```

## Loading and preprocessing the data

Loading the data and converting 'date' string into the proper date format. Also converting 'steps' to numeric format because it store
non-integer types in the future.
```{r echo=TRUE}
activityData <- fread("activity.csv")
activityData[, `:=` (date = as.Date(date), steps = as.numeric(steps))]
```

## What is mean total number of steps taken per day?

Calculating the total number of steps for each day:
```{r echo=TRUE}
dailyTotal <- activityData[, .(stepsSum = sum(steps)), keyby = "date"]
```

Plotting the histogram of total number of steps taken each day:
```{r echo=TRUE}
ggp <- ggplot(data = dailyTotal, aes(x = stepsSum)) +
    geom_histogram(stat = "bin") +
    labs(x = "Total number of steps taken per day")
print(ggp)
```

Now lets calculate the mean total number of steps taken per day:
```{r echo=TRUE}
stepsMean <- mean(dailyTotal$stepsSum, na.rm = TRUE)
```

and now the median total number of steps taken per day:
```{r echo=TRUE}
stepsMedian <- median(dailyTotal$stepsSum, na.rm = TRUE)
```

The mean total number of steps taken per day is `r stepsMean` and the median total number of steps taken per day is `r stepsMedian`.

## What is the average daily activity pattern?

Lets prepare the data:
```{r echo=TRUE}
dailyAverage <- activityData[, .(stepsAverage = mean(steps, na.rm = TRUE)),
                             keyby = .(interval)]
```

And now lets use them to produce the timeseries plot:
```{r echo=TRUE}
ggp <- ggplot(data = dailyAverage, aes(x = interval, y = stepsAverage)) +
    geom_line() +
    labs(y = "Average number of steps", x = "Interval value")
print(ggp)
```

To figure out which interval contains on average the highest number of steps, lets run:
```{r echo=TRUE}
maxInterval <- dailyAverage[stepsAverage == max(dailyAverage$stepsAverage), interval]
```

The 5 minute interval containing the highest average number of total steps is `r maxInterval`.

## Imputing missing values

The total number of missing values:
```{r echo=TRUE}
sum(is.na(activityData$steps))
```

Lets replace the missing data with the average values for the given interval across the dataset:
```{r echo=TRUE}
activityDataI <- merge(activityData, dailyAverage, by = "interval") # merge average values with the original dataset
activityDataI[is.na(steps), steps := stepsAverage] # replace NAs with the average values
```

Plotting the histogram of total number of steps taken each day using the imputed data:
```{r echo=TRUE}
dailyTotalI <- activityDataI[, .(stepsSum = sum(steps)), keyby = "date"]
ggp <- ggplot(data = dailyTotalI, aes(x = stepsSum)) +
    geom_histogram(stat = "bin") +
    labs(x = "Total number of steps taken per day (imputed data)")
print(ggp)
```

Now lets calculate the mean total number of steps taken per day for the imputed data:
```{r echo=TRUE}
stepsMeanI <- mean(dailyTotalI$stepsSum, na.rm = TRUE)
```

and now the median total number of steps taken per day for the imputed data:
```{r echo=TRUE}
stepsMedianI <- median(dailyTotalI$stepsSum, na.rm = TRUE)
```

The mean total number of steps taken per day is `r stepsMean` for the original data and `r stepsMeanI` for the imputed data.
The median total number of steps taken per day is `r stepsMedian` for the original data and `r stepsMedianI` for the imputed data.
As expected, the mean total number of steps hasnt changed. The median total number of steps has increased only slightly.

## Are there differences in activity patterns between weekdays and weekends?

Lets create a new factor variable indicating weekdays and weekends in the imputed dataset:
```{r echo=TRUE}
activityDataI[, weekend := as.factor(weekdays(date) %in% c("Saturday", "Sunday"))]
levels(activityDataI$weekend) <- c("weekday", "weekend")
```

Now lets prepare the avereges data for the plot:
```{r echo=TRUE}
dailyAverageI <- activityDataI[, .(stepsAverage = mean(steps, na.rm = TRUE)),
                               keyby = .(interval, weekend)]
```

And finally lets produce the panel plot:
```{r echo=TRUE}
ggp <- ggplot(data = dailyAverageI, aes(x = interval, y = stepsAverage)) +
    geom_line() +
    facet_grid(weekend ~ .) +
    labs(y = "Average number of steps", x = "Interval value")
print(ggp)
```